---
title: "Belgium Home Search"
output:
  html_document:
    df_print: paged
---

Applying R to automating a home search.

# Location, location, location

As I am preparing for a move, I have been 
spending a lot of time doing research on real estate sites trying to find a rental.  This is compounded by the 
fact that we are able to do very few visits in person due to the sanitary crisis, so we really need to optimize the 
search.  Some factors are easy: we know we want a yard, for example, and we need at least 4 rooms - 
3 for us and the kids and an office for me.  The real estate sites let me set those criteria easily.  
Some are a little bit harder: I'd like a basement, but it's not really a deal breaker if the house doesn't have one
and is otherwise nice.  Then there is the location.  This is easily the most important factor: distance to school, 
work, friends.  This will be my first time having to commute to an office in over 4 years, and even though it 
won't be every day I really don't want it to become a nightmare!  For the school, we have a couple possible schools
picked out (we had a limited pool since we really wanted a bilingual school so the boy can keep practicing his 
English).  So we could be close to either one, but it's really a must - morning minutes are a precious commodity
with 2 kids!  And of course, the friends - it's not that big of a deal if we have to add 5 minute to that trip as 
we won't be doing it every day, but having to drive 45 minutes after a raclette can be trying.

So I wanted a sort of weighted score for the distance which was definitely not an option for the real estate site 
(the best they could offer me was a search by postal code).  And I wanted to be able to consider a home that was
maybe not ideal in other aspects, if it meant we'd be a lot closer to what we wanted (or vice versa, if a dream home
meant adding a few minutes to the commute).  As I have been doing a lot of training with R lately it seemed like the
perfect project to put those skills to the test.  My goal was thus to get a list of homes from the real estate site 
with the characteristic that interested me, geocode them and calculate a distance score for each (according to the 
minimal distance to each of my defined category), then calculate the overall score by using some tweaked factors
(so much for total size, so much for the basement, etc).  Finally plot that against the price to find my dream 
home (or at least, the homes to focus on).

# Sources

 - `pois.csv`, hand-written dataset (points of interest) with:
    - Category  (if there are multiple poi in same category we are going to keep the closest one for each home)
    - Weight (which is category specific, so if there are multiple poi in same category they should have the same weight)
    - Address
    - Lat / Lng (we'll fill those in if missing)
 - `zipcodes_fr_num_new.xls`, postcode data from [BE post](https://www.bpost.be/site/fr/envoyer/adressage/rechercher-un-code-postal)
 - `geocodes.csv`, a list of geocoded towns (from [jedi.be](http://www.jedi.be/blog/2009/04/29/geocoding-belgium-postal-codes/)) - I added the column names
 - `homes.csv`, the homes we want to search - this is obtained by scraping a real estate site with Python.  I describe the process under "Getting the home data".
 
Let's start by including some libraries and getting the base data.


```{r include=FALSE}
library(tidyverse)
library(readxl)
library(sp)
library(httr)
library(xlsx)
# this one will read the .env file where I define my API key
library(dotenv)

pois <- read_delim("pois.csv", delim=";")
geocodes <- read_delim("geocodes.csv", delim=";")
postcodes <- read_excel("zipcodes_num_fr_new.xls", 1)
postcodes$City = iconv(toupper(postcodes$Localité), to="ASCII//TRANSLIT")
geocodes <- geocodes %>%
  left_join(postcodes, "City") %>%
  select(City, PostalCode = "Code postal", Lat, Lng)
rm(postcodes)  # don't need this anymore
```

# Geocoding the POI

Geocoding lets me get the coordinates (Lat and Lng) for each POI.  There is a ggmap package with support for Google
but I used Mapquest instead which has a more permissive license (and this let me practice with httr).  I kept 
the same structure as the one from ggmap though.

Prepare a geocode function.  This will be used for addresses that have a complete street address provided 
(we'll use that both for homes and for POIs)

```{r message=FALSE}
API_KEY = Sys.getenv("MAPQUEST_API_KEY")
COUNTRY = Sys.getenv("COUNTRY")

geocode <- function(location) {
  # vectorize for many locations
  if(length(location) > 1){
    return(map_dfr(as.list(location), geocode))
  }
  NOT_FOUND <- data.frame(Lat=NA, Lng=NA)
  if(is.na(location)) {
    return(NOT_FOUND)
  }
  url = "http://www.mapquestapi.com/geocoding/v1/address"
  r <- GET(url, 
           query = list(key=API_KEY, location=paste(location, COUNTRY, sep=", "), maxResults=1, outFormat="csv")
           )
  if(status_code(r) != 200) {
    warn_for_status(r)
    return(NOT_FOUND)
  }
  parsed <- content(r, "parsed")
  if(nrow(parsed) == 0 || str_ends(parsed$GeocodeQualityCode, "X")) {
    warning(paste("No valid address: ", location))
    return(NOT_FOUND)
  }
  return(parsed %>%  select(Lat, Lng))
}
mutate_geocode <- function(data, location) {
  locs <- data[[deparse(substitute(location))]]
  if(length(locs) == 0) {
    return(data)
  }
  gcdf <- geocode(locs)
  return(bind_cols(data %>% select(-Lat, -Lng), gcdf))
}

# Try it out:
# Some address that exists:
geocode("25 rue des combattants, 1300 Wavre, Belgium")
# One that does not:
geocode("Nonexistent address")
```

Geocode the POIs (if they are missing position)

```{r message=FALSE}
missing <- setdiff(c("Lat", "Lng"), names(pois))
pois[missing] <- NA
if(!all(!is.na(pois$Lat))) {
  pois <- pois %>% mutate_geocode(Address)
}
pois %>% select(-Address)
```


# Getting the home data

I started by getting a list of postal codes with geocodes (`zipcodes_fr_num_new.xls` and `geocodes.csv`).  Then 
figure out all the post codes within a 15 km radius of work.  This let me create a search on the real estate site.
I then used a Python script (not shown here) to scrape the homes.

```{r}
target <- data.matrix(pois[1,] %>% select(Lng, Lat))
zip_points = data.matrix(geocodes %>% select(Lng, Lat))
zip_dists = data.frame(PostalCode=geocodes$PostalCode, Dist=spDistsN1(zip_points, target, TRUE))
zip_dists %>% 
  group_by(PostalCode) %>% 
  summarize(Dist=mean(Dist)) %>% 
  filter(Dist < 15)
```


# Calculate distance score

For each home we are going to:

 - geocode it (if not done yet).  For now we could geocode simply using the postal code?
 - calculate the min distance to each poi category and store it as a column.  Ideally, we'd calculate the distance using mapquest too, so we have actual travel time!  Maybe for the next version.  The road network is dense enough in that area that it is a close enough approximation.
 - calculate the distance score, using the category weights

Now we can read in the home data.

```{r include=FALSE}
homes <- read_csv("homes.csv", col_types="ciicciiddicccccccc")
homes <- drop_na(homes, `Postal code`)    # keep only the ones with a post code
# homes <- filter(homes, Available != "RENTED")
```

I combine it with some address data that was manually collected (by calling the realtors):

```{r}
be = read_excel("Rentals in Belgium.xlsx", 1) %>% select(`Code #`, Address, Available)
homes <- homes %>% 
  left_join(be, by="Code #") %>%
  filter(is.na(Available) || Available != "RENTED")
```


Calculate coordinates for each home that is missing it.  Try to geocode first, for the ones that have an address
provided.  If that fails use the postal code to make a match.

```{r include=FALSE}
homes <- homes %>% 
  mutate(Address = ifelse(is.na(Address) && Street != "", paste(Street, `Postal code`, City, sep=", "), Address))
missing <- is.na(homes$Lat)
geocoded = mutate_geocode(homes[missing,], Address)
homes[missing,]$Lng = geocoded$Lng
homes[missing,]$Lat = geocoded$Lat
missing <- is.na(homes$Lat)
mean_geos <- geocodes %>% 
  group_by(PostalCode) %>% 
  summarise(Lat=mean(Lat), Lng=mean(Lng))
zip_pos <- homes[missing,] %>% 
  left_join(mean_geos, by=c("Postal code" = "PostalCode"), suffix=c(".home", "")) %>%
  select(`Code #`, `Postal code`, Lat, Lng)
homes[missing,]$Lng <- zip_pos$Lng
homes[missing,]$Lat <- zip_pos$Lat
```

Get the points into a matrix (using `data.matrix`), and calculate the distances for each category.
For each category, collect the points into a matrix,
then call `spDists` to obtain a matrix of the distances (each row is a home).
Use `apply` to get the minimum value of each row, and assign the result as the distance for that specific 
category in the home dataframe.
Finally we calculate a total distance, by taking a weighted average of the distances.

```{r}
has_lat <- !is.na(homes$Lat)
Mpoints <- data.matrix(homes[has_lat,] %>% select(Lng, Lat))
score <- rep(0, length(has_lat))
total_weight <- 0
for(cat in unique(pois$Category)) {
  Mtarget <- data.matrix(pois %>% filter(Category == cat) %>% select(Lng, Lat, Weight))
  wt <- Mtarget[1,3]
  Mtarget <- rbind(Mtarget[,-3])
  dists <- spDists(Mpoints, Mtarget, longlat=TRUE)
  min_dists <- apply(dists, 1, min)
  col <- paste("dist.", cat, sep="")
  homes[has_lat, col] <- min_dists
  score <- score + homes[col] * wt
  total_weight <- total_weight + wt
}
homes[has_lat, "dist.Total"] <- score / total_weight
```

To calculate a score we are going to map this into a single number using a simple linear regression, 
it should go from 400 (very good) to 0 (or negative if the house is really far, but we should not even have 
those on the list).  Anything between 0 and 10 is great then it starts getting down more quickly.
We'll set a diminishing return pattern on the size as well, because above 250 sq meter does not help us very much.

```{r}
f <- lm(y ~ x, data=data.frame(x=c(0, 10, 14, 16), y=c(400, 300, 100, 0)))
homes$dist.Score <- predict(f, homes %>% rename(x=dist.Total))
f <- lm(y ~ x, data=data.frame(x=c(0, 250, 350), y=c(0, 250, 270)))
homes$size.Score <- predict(f, homes %>% rename(x=SqMeter))
```


# Calculate total score

Make some flags for the stuff that is in as text, and calculate the new TotalScore column:

```{r}
homes <- homes %>%
  mutate(has_basement =  Basement == "Yes", 
         has_office = Office == "Yes", 
         has_garage = Garage == "Yes",
         has_attic = Attic == "Yes") %>%
  mutate_at(vars(starts_with("has_")), ~replace(., is.na(.), FALSE)) %>%
  mutate(TotalScore = size.Score + 
           Bedrooms * 50 + 
           has_office * 50 + 
           has_garage * 30 + 
           has_basement * 30 +
           has_attic * 15 +
           dist.Score) %>%
  select(-starts_with("has_")) %>%
  arrange(desc(TotalScore))
```


We can now write the results to a new spreadsheet:

```{r}
write.xlsx(x = homes, file="homes_scored.xlsx")
```

Or plot it.  I used plotly to get an interactive chart where I could click on a dot to visit the relevant page,
and have a thumbnail shown when I hover a point (though the images will probably be removed at some point in the
future, since they are pulled directly from the real estate website).

```{r message=FALSE, warning=FALSE}
library(plotly)
library(htmlwidgets)

homes$`Postal code` <- as.character(homes$`Postal code`)
homes$ID <- paste(homes$`Code #`, homes$`Postal code`)

# ggplot(head(homes, 10), aes(x=`Rent`, y=TotalScore, color=ID)) + geom_point()
p_byrent <- plot_ly(head(homes, 25), 
        type="scatter", mode='markers', hoverinfo='text') %>%
  config(displayModeBar = FALSE) %>%
  add_markers(x = ~ Rent, 
              y = ~ TotalScore, 
              text = ~ ID, 
              color = ~ `Postal code`, 
              customdata = ~ paste(Link, Image, sep="|"),  colors="Paired") %>%
  layout(xaxis = list(title = list(text = "Rent €")),
         yaxis = list(title = list(text = "Score")),
         title = "Score by Rent") %>%
  # image tooltips from https://plotly-r.com/supplying-custom-data.html#fig:tooltip-image
  htmlwidgets::onRender(readLines("tooltip-images.js"))
p_bydist <- plot_ly(head(homes, 25), 
        type="scatter", mode='markers', hoverinfo='text') %>%
  config(displayModeBar = FALSE) %>%
  add_markers(x = ~ dist.School, 
              y = ~ dist.Work, 
              text = ~ ID, 
              color = ~ `Postal code`, 
              customdata = ~ paste(Link, Image, sep="|"),  colors="Paired") %>%
  layout(xaxis = list(title = list(text = "School km")),
         yaxis = list(title = list(text = "Work km")),
         title = "Distances") %>%  
  # image tooltips from https://plotly-r.com/supplying-custom-data.html#fig:tooltip-image
  htmlwidgets::onRender(readLines("tooltip-images.js"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
p_byrent
p_bydist
```

And I found my dream home!  Well, not really, because the top one has some other issues, but at least this gets us closer.

